% This is a template for use with the MSU Thesis class
%
% Class options: 
% PhD for dissertations;
% MA for Master of Arts 
% MS for Master of Science
% MAT for Master of Arts for Teachers  
% MBA for Master of Business Administration  
% MFA for Master of Fine Arts  
% MIPS for Master of International Planning Studies  
% MHRL for Master of Human Resources and Labor Relations  
% MMus for Master of Music  
% MSN for Master of Science in Nursing  
% MPP for Master of Public Policy  
% MSW for Master of Social Work  
% MURP for Master in Urban and Regional Planning  
%
% Default is PhD
%
%
% This template has everything in the right order.
% Just add real content and you're done!
%
\documentclass[lscape]{msu-thesis}
%
% for a prettier, but possibly non-compliant table of contents use the [mixedtoc] option
% for a plain table of contents use the [plaintoc] option
% for a horrendous looking, but possibly required table of contents, use the [boldtoc] option
%
% If you have large tables/figures that need to be in landscape mode, add the [lscape] option

% This is standard fontenc/inputenc for pdflatex
% If you use LuaLaTeX or XeLaTeX you should replace this with the fontspec package
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
%
% If the thesis office requires Times, we'll give them Times
% You can experiment with other font packages here if you like.
% If you are using XeLaTeX or LuaLaTeX load the Times or Times New Roman font with \setmainfont
\usepackage{mathptmx} 
\graphicspath{{images/}}
%
% Load any extra packages here
%
% You must specify the title of your thesis, your name, the field of study (not department), and the year
\title{Identification of candidate genes and isoforms associated with
genetic resistance to Marek's Disease from RNA-Seq data}
\author{Likit Preeyanon}
\fieldofstudy{Microbiology and Molecular Genetics} % This should be in sentence case
\date{2014}

% If you want a dedication page, specify the text of the dedication here and uncomment the next command.
%
\dedication{To scientists who encourage open science.}
%
\begin{document}

% All the stuff before your actual chapters is called the front matter
\frontmatter
% First make the title page
\maketitlepage
% Next make the abstract
\begin{abstract}

Marek's disease (MD) in chickens is characterized by T cell
lymphomas caused by the Marek's disease virus (MDV), an
$\alpha$-herpesvirus. MD is a major economical problem for
the poultry industry as it causes approximately \$2 billion
in worldwide losses annually. Although vaccination has been
effective at preventing tumor formation, it has not been
able to prevent MDV infection or replication.  Consequently,
more virulent field strains have emerged over the past
decades following the introduction of new vaccines. Poor
practices of vaccination and incomplete immunity have been
speculated to play a role in driving the evolution of the
virus with greater virulence. Therefore, it is critically
important to develop more sustainable control measures to
the disease in the long run.

Development of genetically resistant chickens has been an
alternative approach to control the virus and a number of studies
have been conducted to identify specific genes that contribute to
MD resistance. The major histocompatibility (MHC) locus has been
found to be strongly associated with resistance or susceptibility
to MD, and several alleles have been well characterized.  Non-MHC
genes also play a major role in resistance to MD. Two inbred
lines (line 6 and line 7) maintained at Avian and Oncology
Laboratory share the same MHC allele (B\textsuperscript{2}), yet
line 6 is resistant and line 7 is susceptible to MD,
respectively.  These two lines have been used as a model to study
non-MHC genes that contribute to resistance and susceptibility to
the disease.

To identify non-MHC genes contributing resistance to MD, a
computational pipeline was developed to integrate gene models
from Ensembl, {\em de novo} assembly, and reference-based
assembly (Cufflinks) of sequencing reads to construct a more
complete set of gene models that include more complete
untranslated regions (UTRs) and isoforms predicted from RNA-Seq
data. The results from expression analysis suggest that the
immune response in line 7 is more active at the early stage of
infection (4 days post-infection) compared to line 6.
Differentially expressed genes are enriched in pathways involved
in both the innate and the adaptive immune response in line 7,
whereas, only genes involved in the innate immune response are
significantly enriched in line 6. Due to the cell-associated
nature of MDV and the current model of MDV infection, the virus
is thought to transfer from B cells and antigen presenting cells
(APCs) to activated T cells during the lytic infection.
Therefore, repressed or delayed activation of the adaptive immune
response in line 6 may be a key mechanism conferring MD
resistance.

Investigation of differential exon usage suggests that genes
involved in the cytoskeleton pathway may play a role in
repressing the activation of the adaptive immune response.
For instance, the {\em ITGB2} gene encodes integrin $\beta$2, a component
of several molecules including the lymphocyte
function-associated antigen 1 (LFA-1). LFA-1 is exclusively
expressed on the surface of leukocytes and plays an
important role in cell-to-cell contact and antigen
presentation.  It could be speculated that an alternative
isoform of {\em ITGB2} affects a function of LFA-1 and prevents T
cells from being activated by APCs or B cells resulting in
the delayed activation of the adaptive immune response or
the lower number of activated T cells, the target of MDV.

The results from this study show that many genes not identified
as differentially expressed at a gene level are differentially
expressed at an isoform level; therefore, they will not be
identified by gene expression analysis alone. Using the pipeline
developed in this study, one can iteratively incorporate ENSEMBL
models and RNA-seq data to construct better gene models that
include genes and isoforms expressed in all samples and perform
differential gene and isoform expression analysis to identify
genes and isoforms that are responsible for resistance to MD.

Although functions of most isoforms are not fully annotated,
we have shown that methods, such as protein prediction and
pathway analysis, can be used to predict the putative
functions of the isoforms and their potential roles in MD
resistance, which could open up a new direction for MD
research. Moreover, prediction of causative {\em cis}-regulatory
elements in those genes will lead to identification of
precise genetic factors contributing to MD resistance.

\end{abstract}

% Force a newpage
\clearpage
% Make the copyright page. The Graduate School ridiculously prohibits you
% from having a copyright page unless you pay ProQuest to register the copyright.
% This should be illegal, but I didn't make up the rule.

% \makecopyrightpage

% If you have a dedication page, uncomment the next command to print the dedication page
\makededicationpage
%
\clearpage
% Your Acknowledgements are formatted like a chapter, but with no number
\chapter*{Acknowledgements}
\DoubleSpacing % Acknowledgements should be double spaced
\input{ack}
\clearpage
% We need to turn single spacing back on for the contents/figures/tables lists
\SingleSpacing
\tableofcontents* % table of contents will not be listed in the TOC
\clearpage
\listoftables % comment this out if you have no tables
\clearpage
\listoffigures % comment this out if you have no figures
%
% If you have a list of abbreviations/symbols it would go here preceded by a \clearpage
%
\mainmatter
%
% The next line removes the dots in chapter headings in the TOC
% May violate thesis office rules
%\addtocontents{toc}{\protect\renewcommand{\protect\cftchapterdotsep} {\cftnodots}}

\input{background}
\input{chaptertwo}
\input{chapterthree}
\input{chapterfour}
\input{conc}

%
% If you have pages that must appear in landscape mode, use the [lscape] option
% and enclose the pages in a {landscape} environment.
%\clearpage\pagestyle{lscape} % first clear the page and change the pagestyle
%\begin{landscape}
%
% your landscape table(s) or figure(s) here
%
%\end{landscape}
%\pagestyle{plain} % remember to change the pagestyle back to plain
%
%
% If you have appendices, they would go here.  
% Comment these lines out if you don't
% If you have more than one appendix, uncomment the next line and comment out the \appendix line
%
%\appendices
% \makeappendixcover
\clearpage
\appendix
\chapter*{\ }
\begin{center}
\textbf{APPENDIX}

\textbf{GLOSSARY}
\end{center}
\input{gloss}
%
\backmatter
% The next lines add the dots back into the References/Bibliography heading
% of the TOC.  Only uncomment this if you need to put the dots back in having
% removed them for Chapter headings.
%
%\addtocontents{toc}{%
%   \protect\renewcommand{\protect\cftchapterdotsep} {\cftdotsep}}
%
\makebibliographycover % make the bibliography cover page
% Bibliography can be single spaced
%
\SingleSpacing
% Your bibliography command here (e.g. \bibliography{your-bib-file}) if using natbib
\bibliographystyle{abbrv}
\bibliography{dis}
\end{document}

